On the Gravitational stability of gravito-turbulent accretion disks

### 2016-03-23 paper/model3d.tex
revising, point 4

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

\subsubsection{Beta cooling}\label{beta_cool_model}
In our beta cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}. 
\end{align}
Recall  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter. Here, 
\begin{align}\label{tirr_def}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant 
fraction $\theta$ of the equilibrium temperature
$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
for example, stellar or background irradiation. Note that $\theta$  
should only be interpreted as an irradiation parameter when it is 
defined through Eq. \ref{tirr_def} and 
Eq. \ref{beta_cool} adopted as the cooling function.

For beta cooling given by Eq. \ref{beta_cool} we assume an 
equilibrium polytropic relation 
\begin{align}\label{poly_vert} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(z=0)$ is the equilibrium mid-plane density. 
The vertical structure is first obtained from 
Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
then inserted into 
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right). 
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align}

>>>>

\section{Basic equations}\label{basic}
We consider a 3D, self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
to the radial, azimuthal and vertical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field. 
We assume an ideal gas so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}\label{ideal_eos}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
{\bf For simplicity we refer to a single adiabatic index as that
  appears in the equation of state for both 2D and 3D disk models 
  \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
                             %for the same 
                             %disk in 2D 
%johnson: specify 3D gamma, then translate into 2D gamma because disk
%model is 2D 
}
The gas 
gravitational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho. 
\end{align}

@@

\subsubsection{Beta cooling}\label{beta_cool_model}
In our beta cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}.
\end{align}
Recall  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter. 
{\bf Here, $T_\mathrm{irr}$}
%Here, 
%\begin{align}\label{tirr_def}
%  \tirr = \theta T_\mathrm{eq}
%\end{align}
is a reference temperature field. 
%parameterized as a 
%fraction $\theta$ of the equilibrium temperature field 
%$T_\mathrm{eq}(z)$. 
Physically, $\tirr$ {\bf may} the floor temperature set by, 
for example, stellar or background irradiation. {\bf

}

{\bf
Note that $\theta$  
should only be interpreted as an irradiation parameter when it is 
defined through Eq. \ref{tirr_def} and 
Eq. \ref{beta_cool} adopted as the cooling function. 
}

For beta cooling given by Eq. \ref{beta_cool} we assume an 
equilibrium polytropic relation 
\begin{align}\label{poly_vert} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(z=0)$ is the equilibrium mid-plane density. {\bf Note that
  $\Gamma$ is only relevant 
  to the 3D problem as it determines the disk vertical structure.} 
The vertical structure is first obtained from 
Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
then inserted into 
Eq. \ref{thermal_eq} to infer the required viscosity profile for
thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right). 
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align}

### 2016-03-23 paper/paper3d.tex
begin revision
updated aknowledge list
revising, point 4

\begin{abstract}
  Low mass, self-gravitating accretion disks admit quasi-steady,
  `gravito-turbulent' states in which cooling balances turbulent viscous 
  heating. However, numerical simulations show that gravito-turbulence
  cannot be sustained beyond dynamical timescales when the cooling
  rate or corresponding turbulent viscosity is too large. The result is disk fragmentation.         
  We motivate and quantify an interpretation of disk 
  fragmentation as the inability to maintain gravito-turbulence due to  
  formal secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
%We analyze the gravitational stability of accretion disks 
%including these non-ideal effects explicitly.    
%We derive a general dispersion relation for linear waves in a 
%self-gravitating, viscous, non-adiabatic disk with internal heating,
%external irradiation, and cooling. 
We analyze the gravitational stability of viscous, non-adiabatic
accretion disks with internal heating, external irradiation, and cooling.  
We consider parameterized cooling functions in two and three dimensions,
as well as 
radiative diffusion in 3D. We show that 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable, although interesting limits appear 
for unstable modes with lengthscales on the order of the disk thickness.   
%when we examine the growth 
%of modes of order the disk scale height. 
We apply this new linear theory to 
models of protoplanetary disks subject to
gravito-turbulence modeled as an effective 
viscosity, and cooling regulated by dust opacity. 
We find that viscosity renders the disk beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity, and that the observed transition in behavior from 
gravito-turbulent to fragmenting may derive from  
instability of the gravito-turbulent state itself.
\end{abstract}

\section{Introduction}\label{intro} 
Understanding the gravitational stability of rotating disks    
is central to many astrophysical problems. In the context of 
gaseous protostellar or protoplanetary disks (PPDs), gravitational instability (GI)
has two applications. It can provide gravitational torques to
transport angular momentum outwards and thus enable mass accretion
\citep{armitage10,turner14}. GI may also lead to disk fragmentation,
which has been invoked to explain the formation of stellar/sub-stellar companions or 
giant planets 
at large radii \citep{boss97,kratter06,stam09,helled14} %% KMK need an original citation, not just review.
Studying these non-linear phenomena requires
direct numerical simulations. Nevertheless, physical insight can be    
obtained through analytical modeling.

@@

The goal of this work is to generalize the analytic treatment of disk
GI, by including cooling and viscosity, to allow 
a self-consistent discussion of GI in realistic accretion disk
models.  
%Specifically, we will examine how  cooling and viscosity can
%lead to GI even when $Q>1$, by removing pressure and rotational
%support, respectively.  
We first review in \S\ref{cool_gi} and \S\ref{visc_gi}, respectively,  
how cooling or viscosity can lead to GI even when $Q>1$, by removing
pressure or rotational support. We then 
discuss in \S\ref{frag_intro} how these effects may relate   
to the transition from self-regulated GI to fragmentation seen in
numerical simulations. The rest of this paper is laid out in
\S\ref{plan}.

@@

Consider an initially laminar, $Q\gg1$ disk, without external 
heating, as it cools. The disk temperature will decline
until $Q=O(1)$, whence non-axisymmetric modes grow and heat the disk
through the dissipation of spiral shocks \citep{cossins09}. 
%new rafikov 16  
This setup permits a quasi-steady, turbulent state with
$Q =O(1)$ in which cooling is balanced by shock heating to maintain
thermal equilibrium \citep{gammie01}. 
If one represents this turbulent heating by a viscous heating, then
the $\alpha$ and $\beta$ parameters defined above will  
be inversely related (see, e.g. Eq. \ref{alpha_beta_relation}).

@@

>>>>

\begin{abstract}
  Low mass, self-gravitating accretion disks admit quasi-steady,
  `gravito-turbulent' states in which cooling balances turbulent viscous 
  heating. However, numerical simulations show that gravito-turbulence
  cannot be sustained beyond dynamical timescales when the cooling
  rate or corresponding turbulent viscosity is too large. The result is disk fragmentation.         
  We motivate and quantify an interpretation of disk 
  fragmentation as the inability to maintain gravito-turbulence due to  
  formal secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
%We analyze the gravitational stability of accretion disks 
%including these non-ideal effects explicitly.    
%We derive a general dispersion relation for linear waves in a 
%self-gravitating, viscous, non-adiabatic disk with internal heating,
%external irradiation, and cooling. 
We analyze the {\bf axisymmetric} gravitational stability of viscous, non-adiabatic
accretion disks with internal heating, external irradiation, and
cooling {\bf in the shearing box approximation}.   
We consider parameterized cooling functions in {\bf 2D} and {\bf 3D}
disks, %two and three dimensions,  
as well as   
radiative diffusion in 3D. We show that 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable, although interesting limits appear 
for unstable modes with lengthscales on the order of the disk thickness.   
%when we examine the growth 
%of modes of order the disk scale height. 
We apply this new linear theory to 
models of protoplanetary disks subject to
gravito-turbulence modeled as an effective 
viscosity, and cooling regulated by dust opacity. 
We find that viscosity renders the disk beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity, and that the observed transition in behavior from 
gravito-turbulent to fragmenting may derive from  
instability of the gravito-turbulent state itself.
\end{abstract}

\section{Introduction}\label{intro} 
Understanding the gravitational stability of rotating disks    
is central to many astrophysical problems {\bf \citep{kratter16}}. In the context of 
gaseous protostellar or protoplanetary disks (PPDs), gravitational instability (GI)
has two applications. It can provide gravitational torques to
transport angular momentum outwards and thus enable mass accretion
\citep{armitage10,turner14}. GI may also lead to disk fragmentation,
which has been invoked to explain the formation of stellar/sub-stellar companions or 
giant planets 
at large radii \citep{boss97,kratter06,stam09,helled14} %% KMK need an original citation, not just review.
Studying these non-linear phenomena requires
direct numerical simulations. Nevertheless, physical insight can be    
obtained through analytical modeling.

@@

The goal of this work is to generalize the analytic treatment of disk
GI, by including cooling and viscosity, to allow 
a self-consistent discussion of GI in realistic accretion disk
models.  
%Specifically, we will examine how  cooling and viscosity can
%lead to GI even when $Q>1$, by removing pressure and rotational
%support, respectively.  
We first review in \S\ref{cool_gi} and \S\ref{visc_gi}, respectively,  
how cooling or viscosity can lead to GI even when $Q>1$, by removing
pressure or rotational support. We then 
discuss in \S\ref{frag_intro}{\bf---\S\ref{gi_of_gi}} how these
effects may relate to the transition from self-regulated GI to
fragmentation seen in numerical simulations. The rest of this paper is
laid out in \S\ref{plan}.

@@

Consider an initially laminar, $Q\gg1$ disk, without external 
heating, as it cools. The disk temperature will decline
until $Q=O(1)$, whence non-axisymmetric modes grow and heat the disk
through the dissipation of spiral shocks \citep{cossins09}. 
%new rafikov 16  
This setup permits a quasi-steady, turbulent state with
$Q =O(1)$ in which cooling is balanced by shock heating to maintain
thermal equilibrium \citep{gammie01}. 
%If one represents this turbulent heating and transport {\bf by a viscosity} 
%heating by a viscous heating

{\bf 
  Global disk simulations \citep[e.g.][]{lodato04} show
  that the transport and 
  heating associated with this gravito-turbulence may be described as a
  local viscous process provided that the disk-to-star mass ratio is
  small ($\lesssim0.25$) and the disk is thin (aspect-ratio $\lesssim
  0.1$).  
}
{\bf In this case} the $\alpha$ and $\beta$ parameters defined above
will be inversely related (see, 
e.g. Eq. \ref{alpha_beta_relation}). {\bf Recent 
  vertically-extended shearing box simulations 
  also confirm this relation in 3D disks \citep{shi14}.}

@@

{\bf 
\subsection{Instability of the gravito-turbulent state}\label{gi_of_gi}
  In this work, we formally regard the gravito-turbulent
  disk described above as an equilibrium state, to which we apply    
  standard linear stability analysis. Then by `perturbations' we mean   
  deviations away from this gravito-turbulent 
  basic state\footnote{This should not be confused with 
    flucuations with respect to the 
    laminar disk that maintain the underlying
    gravito-turbulence.}. 
  We assume that instabilities defined this way  
  signifies disk fragmentation, since numerical simulations  
  show that the non-existance of steady gravito-turbulence imply    
  fragmentation.   
}
